---
title: "chapter5.Rmd"
author: "Ahi"
date: "2 joulukuuta 2017"
output: html_document
---

```{r}
library(GGally)
library(dplyr)
library(corrplot)
```

# 1. Loading data into R

In this exercise we use a real dataset developed by the UN to measure the humand development of countries in the world. 
We now explore the struture and the dimensions of the data.

```{r}
human2 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep  =",", header = T)

str(human2)
dim(human2)
summary(human2)
```

Now we look at the normal disctributions, correlationshs, and a graphical overview.  
```{r}
ggpairs(human2)
```
As an example to explore the graph above, We can see that expected years of schooling (Edu.Exp) has almost a normal distrbution. We can also understand that there is quite a strong positive relationship between female/male ratio of secondary education (Edu2.Fm) and life expectancy at birth (Life.Exp).  

Let us also look at the correlations of the variables.
```{r}
cor(human2)
```

```{r}
cor(human2) %>% corrplot
```



